\propertitle

### 2019-07-22 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-21T17:05:52+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the neurosciences the maximum-entropy method is often applied to
questions of sufficient statistics. In this section we discuss how our
proposed application bears on this kind of questions. For concreteness we
use an example with specific statistics: first and second moments
(equivalent to mean and correlations), versus first-to-fourth moments. But
our discussion holds for any  sets of statistics.

The question is as follows. We want to assess our degree of belief about
the frequencies of the activities of the sample, or about the sample
activity in a new time bin. For this assessment we should use all measured
data we have. But it sometimes happens that dropping part of the data --
for example, the measured third- and higher-order moments -- leaves our
assessment almost unchanged: the dropped data are \emph{informationally
  irrelevant}, or almost so. The remaining data -- for example, first and
second moments -- are \emph{informationally sufficient}. This informational
quasi-sufficiency is interesting because it may hint at peculiar biological
or physical mechanisms or dependencies.

We can approximately check how much some data are irrelevant by simply
dropping them from the conditional of our degree of belief, and see if the
latter changes appreciably. When our degree of belief is built with the
maximum-entropy method, we simply drop some of the data used as constraints
and see how the resulting distribution changes. For example, we may only
retain the first two moments (means and correlations) or the first four
moments. In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from.

@@

The method introduced in the present work reveals an issue with this
application at the sample level, however. The issue doesn't come from the method
itself but from a general fact of probability theory. If a probability
distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically it corresponds to the general
impossibility of solving a system of $n$ equations in $m$ unknowns when
$n>m$ \citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some specific, casually
recorded sample. Thus, paradoxically, maximum-entropy methods applied at
the sample level give an answer opposite to the one we might think they're
giving.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. A correct answer can only be given case by case;
\fig~\ref{fig:comparisons_marginals} gives a graphical answer to this
question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{mypurpleblue}{\textbf{------}}). The lower plot
shows the maximum-entropy distributions for the sample, from two moments
(\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{mypurpleblue}{$\bm{\square}$}), and the distributions obtained by
marginalizing the full-network distribution to the sample size, from two
moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$}) and from four
moments (\textcolor{mypurpleblue}{$\bm{\blacksquare}$}). The measured frequency
distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also shown. We
note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{mypurpleblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.

@@

@@

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-22T11:30:14+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In the neurosciences the maximum-entropy method is often applied to
questions of sufficient statistics; for example: do pairwise correlations
in the neuronal activity provide the same information as the full frequency
distribution of activities? In this section we discuss how our proposed
application bears on this kind of questions. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our discussion
% holds for any sets of statistics.

Let's make the question more precise first. We want to assess our degree of
belief about the frequencies of the activities of the sample, or about the
activity of the sample in a new time bin. For this assessment we should use
all measured data we have. But it sometimes happens that dropping part of
the data -- for example, the measured third- and higher-order moments --
leaves our assessment almost unchanged: the dropped data are
\emph{informationally irrelevant}, or almost so. The remaining data -- for
example, first and second moments -- are then \emph{informationally
  sufficient}. This informational quasi-sufficiency is interesting because
it may hint at peculiar biological  mechanisms or dependencies.

We can approximately check how much a part of the data is irrelevant by
simply dropping it from the conditional of our degree of belief, and see
if the latter changes appreciably. When our degree of belief is built with
the maximum-entropy method, we simply drop part of the data used as
constraints and see how the resulting distribution changes. For example, we
may only retain the first two moments (means and correlations) or the first
four moments.

@@

In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from. The method introduced in the present work, however,
reveals an issue with this application at the sample level. The issue
doesn't come from the method itself but from a general fact of probability
theory. If a probability distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically this impossibility translates into a
system of $n$ independent equations in $m$ unknowns with $n>m$
\citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some specific, casually
recorded sample. Thus, paradoxically, maximum-entropy methods applied
\emph{at the sample level} give an answer opposite to the one we might
think they're giving.

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
%comparison_ME_sample_full_levels.nb
But does the application at the full-network level lead to appreciably
different results from that at the sample level? after all we're interested
in an approximate informational sufficiency, not in an analytically exact
one. A correct answer can only be given case by case.
Figure~\ref{fig:comparisons_marginals} gives a graphical answer to this
question in the case of our first data set. The upper plot shows the
maximum-entropy distributions for a full network of $10\,000$ neurons,
constructed from two moments (\textcolor{mygreen}{\textbf{--~--~--}}) and
from four moments (\textcolor{mypurpleblue}{\textbf{------}}). The lower
plot shows the maximum-entropy distributions for the sample, from two
moments (\textcolor{mygreen}{$\bm{\triangledown}$}) and from four moments
(\textcolor{mypurpleblue}{$\bm{\square}$}), and the distributions obtained
by marginalizing the full-network distribution to the sample size, from two
moments (\textcolor{mygreen}{$\bm{\blacktriangledown}$}) and from four
moments (\textcolor{mypurpleblue}{$\bm{\blacksquare}$}). The measured
frequency distribution (\textcolor{myred}{\LARGE$\bm{\bullet}$}) is also
shown. We note the following:
\begin{itemize}
\item The application to the full-network (upper plot) leads to two
  completely different distributions (even different number of modes);
  clearly the two sets of statistics are not even approximately equivalent
  for inferential purposes.
\item The application at the sample level leads to deceivingly similar
  distributions (\textcolor{mygreen}{$\bm{\triangledown}$} and
  \textcolor{mypurpleblue}{$\bm{\square}$}), which can only be distinguished on a
  logarithmic scale. This could lead to the erroneous conclusion that the
  two statistics are approximately equivalent.
\item The difference between marginals from the full-network distribution
  and the distributions obtained at the sample level (filled vs empty
  markers) is larger than the difference between different statistics
  (triangles vs squares).
\item The marginals of the application to the full network are closer to
  the measured frequencies than the distributions obtained at the sample
  level (although this closeness is not a valid criterion for their
  goodness).
\end{itemize}
Therefore, the maximum-entropy application at the sample level and at the
full-network level lead to different results; the latter is not only more
meaningful but also superior because it shows clearly the different
informational sufficiency of the two sets of statistics.

@@

In the study or use of the frequency distribution obtained with the
procedure here presented we must take into account two important points.

The first point is that there are many possible frequency distributions
which we believe, to different degrees, could be the true one that happened
during the recording. The one given by our procedure is simply the one with
the largest degree of belief, the mode of the belief distribution. The
space of possible frequency distributions has many dimensions, however --
thousands or tens of thousands. We must remember that belief distributions
in high dimensions have counter-intuitive properties. For example, the mode
or mean can have \emph{atypical} features when compared with the features
of most other points of the space. The mode and mean can also be very
different from each other.

The question, then, is \emph{which features of the maximum-entropy
  frequency distribution are typical of the majority of plausible frequency
  distributions?} We can only answer for sure by using the full-fledged
probability calculus. A more complete study (in preparation) with the first
data set reveals that most of the plausible frequency distributions have
these important features in common with the maximum-entropy one:
\begin{itemize}
\item all activities $\yA \lesssim 0.05\%$ have practically zero frequencies;
\item there are two regions of activity levels with high frequencies,
  roughly separated by a trough of lower frequencies. The frequencies of
  the region on the left ($\yA \lesssim 0.018\%$) are higher than those of
  the region on the right ($\yA \gtrsim 0.018\%$).
\end{itemize}
But there are also differences. For example, plausible frequency
distributions may have three or four modes instead of just two; and these
modes are higher than of the the maximum-entropy distribution.

The second point is that our degrees of belief about the frequency
distribution for the full network depend not only on the measured data in
the sample, but also on our pre-data beliefs about the distribution. Which
assumptions lie behind the maximum-entropy approximation?

@@

We have presented a procedure to construct the most plausible frequency
distribution of population-averaged activities of a network of neurons,
given the recording about a small sample thereof. This procedure combines
the maximum-entropy method and basic identities from sampling theory. From
the application to two real data sets we saw that the frequency
distributions obtained with our procedure can have features very different
from the one measured in the sample, such as multiple modes. This procedure
can also be used with moment constraints of different order -- means,
population-averaged pairwise correlations, or higher-order correlations --
thus giving an approximate assessment of the informational sufficiency of
specific subsets of moments. In fact, we saw that the application of
maximum-entropy only at the sample level leads to misleading results about
this kind of sufficiency questions.

Let's
